---
title: "Weekends Clean the Air of Traffic Pollution, but Not of Ozone"
subtitle: "Weekly cycles in US urban air quality, 2000-2023"
author:
  - "Soham"
date: "July 28, 2026"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    embed-resources: true
    fig-width: 9
    fig-height: 5
    code-fold: false
execute:
  echo: false
  warning: false
  message: false
  cache: false
editor: visual
---

```{r}
#| label: setup
library(tidyverse)
library(knitr)

source("R/load_data.R")
source("R/models.R")
source("R/plots.R")
source("R/summaries.R")

# Analysis settings. Nothing downstream hardcodes these values.
n_locations <- 20
outcome_columns <- c(
  "Nitrogen dioxide (NO2)" = "no2_mean",
  "Ozone (O3)" = "o3_mean"
)

analysis_bundle <- build_analysis_bundle(
  path = resolve_data_path(),
  n_locations = n_locations
)

analysis_data <- analysis_bundle$data
```

```{r}
#| label: fit-models
weekend_models <- fit_all_models(
  analysis_data, outcomes = outcome_columns, weekly_term = "is_weekend"
)

day_models <- fit_all_models(
  analysis_data, outcomes = outcome_columns, weekly_term = "day_of_week"
)

weekend_effects <- summarize_weekly_effects(weekend_models, "is_weekend")
day_effects <- summarize_weekly_effects(day_models, "day_of_week")
```

# Introduction

## The data

This report uses the US Air Quality Metrics dataset, which compiles daily
readings from Environmental Protection Agency monitoring stations between
January 2000 and September 2023. The raw file contains 665,414 monitor-days
across 21 measured columns, covering 48 states and 150 cities. For each
monitor-day it records the daily mean, the daily maximum, the hour of that
maximum, and the Air Quality Index for four pollutants: nitrogen dioxide
(NO~2~), ozone (O~3~), carbon monoxide (CO), and sulphur dioxide (SO~2~).

```{r}
#| label: tbl-summary
#| tbl-cap: |
#|   Summary statistics for the two outcome variables, measured across the
#|   analysis sample.
analysis_data |>
  summarize_outcomes(outcomes = outcome_columns) |>
  kable(digits = c(0, 0, 3, 3, 3, 3), format.args = list(big.mark = ","))
```

After cleaning and restricting to the `r n_locations` best-covered
monitoring locations, the analysis sample contains
`r format(nrow(analysis_data), big.mark = ",")` monitor-days spanning
`r min(year(analysis_data$date))` to `r max(year(analysis_data$date))`.
Nitrogen dioxide, the primary outcome, averages
`r round(mean(analysis_data$no2_mean), 1)` ppb with a standard deviation of
`r round(sd(analysis_data$no2_mean), 1)` ppb.

## Code book

@tbl-codebook defines every variable used in the analysis, including the
derived calendar variables.

```{r}
#| label: tbl-codebook
#| tbl-cap: "Data dictionary for all variables used in the analysis."
build_code_book() |>
  kable()
```

## Variables omitted, and observations dropped

The raw file carries several columns that this analysis does not use.
@tbl-omitted records each omission and the reason for it.

```{r}
#| label: tbl-omitted
#| tbl-cap: "Raw variables excluded from the analysis, with justification."
build_omitted_variables() |>
  kable()
```

Two cleaning rules remove observations, and @tbl-audit reports the cost of
each. First, monitors occasionally report zero or slightly negative
concentrations when true levels sit near the instrument detection limit.
These readings are physically impossible and undefined under the log
transformation used below, so they are dropped. Second, a small number of
monitors carry the placeholder city label "Not in a city" and cannot be tied
to an urban area, so they are dropped as well. Together these rules remove
roughly 9% of the raw file, and neither rule depends on the day of the week,
so neither should bias a weekday-versus-weekend comparison.

```{r}
#| label: tbl-audit
#| tbl-cap: "Observations remaining after each cleaning step."
analysis_bundle$audit |>
  kable(
    col.names = c("Cleaning stage", "Rows remaining", "Rows dropped"),
    format.args = list(big.mark = ",")
  )
```

The analysis then restricts attention to the `r n_locations` locations with
the most complete records. This keeps the panel balanced across the full
24-year window and concentrates on dense urban monitors, where any traffic
signal should be strongest. The cost of this choice is discussed in the
critique.

## Motivation

Nitrogen dioxide in cities comes overwhelmingly from combustion, and in
particular from motor vehicle exhaust. Commuting traffic follows a sharp
weekly rhythm, so if vehicle emissions drive urban NO~2~, concentrations
should fall on Saturdays and Sundays. This "weekend effect" is a natural
experiment that repeats every seven days for 24 years, which makes it an
unusually clean way to look for a human fingerprint in environmental data.

Ozone is the interesting complication. Ground-level ozone is not emitted by
cars at all; it forms in the atmosphere through sunlight-driven reactions
involving nitrogen oxides and volatile organic compounds. Atmospheric
chemistry offers a specific reason to expect ozone to behave differently:
freshly emitted nitric oxide destroys ozone directly in a reaction known as
titration. Where that scavenging effect dominates, cutting traffic emissions
should *raise* ozone rather than lower it. If both patterns appear in the
same data, the weekly cycle becomes a way to see two different chemical
regimes at once.

## Research question and hypotheses

> **Question.** Do weekly cycles in human activity leave a measurable
> fingerprint on urban air quality, and do nitrogen dioxide and ozone
> respond in the same direction?

Writing $\beta$ for the coefficient on a weekend indicator in a model of log
concentration, the analysis tests two directional hypotheses.

For nitrogen dioxide, the traffic-marker hypothesis:

$$
H_0: \beta_{\text{NO}_2} = 0 \qquad
\text{versus} \qquad
H_A: \beta_{\text{NO}_2} < 0
$$

For ozone, the titration hypothesis:

$$
H_0: \beta_{\text{O}_3} = 0 \qquad
\text{versus} \qquad
H_A: \beta_{\text{O}_3} > 0
$$

We expect to reject both null hypotheses, and specifically we expect the two
pollutants to move in opposite directions: lower NO~2~ on weekends, higher
O~3~ on weekends.

# Methods

## Model

Both outcomes are modelled with ordinary least squares on the natural log
scale:

$$
\log(y_{it}) = \beta \cdot \text{weekend}_t + \gamma_{m(t)} +
\delta_{y(t)} + \alpha_i + \varepsilon_{it}
$$

for monitor location $i$ on date $t$. The log transformation serves two
purposes. Concentrations are strictly positive and strongly right-skewed, as
@fig-logscale shows, and logging both stabilises the spread and makes the
coefficient directly interpretable: $100 \times (e^{\beta} - 1)$ is the
approximate percentage difference between weekends and weekdays.

The three control terms absorb the largest competing sources of variation.
Month fixed effects ($\gamma$) absorb the strong seasonal cycles visible in
both pollutants; ozone in particular peaks in summer because its formation
is sunlight-driven. Year fixed effects ($\delta$) absorb the long-run
downward trend in emissions shown in @fig-trend. Location fixed effects
($\alpha$) absorb persistent differences between cities, so the weekend
coefficient is identified from variation *within* a monitoring site rather
than by comparing clean cities to dirty ones.

A second specification replaces the binary weekend indicator with a
seven-level day-of-week factor, using Wednesday as the reference. This
produces the main figure and shows whether the effect is a genuine weekend
pattern or an artefact of one unusual day.

## Assumptions

Ordinary least squares requires linearity in the parameters, independent
errors, constant error variance, and, for exact small-sample inference,
normally distributed errors. Linearity is imposed on the log scale, and the
fixed-effect structure means every predictor is categorical, so no
functional form is assumed for the controls. The remaining three assumptions
are checked in @sec-validity, where two of them turn out to be imperfect.

## Software

The analysis uses R with the `tidyverse` collection: `readr` for input,
`dplyr` and `tidyr` for manipulation, `lubridate` for calendar variables,
`purrr` to fit and summarise both models without repeating code, and
`ggplot2` for figures. Models are fitted with base R's `lm()`, and estimates
are extracted with `summary()` and `confint()`. Tables are rendered with
`knitr::kable()`. All data preparation, model fitting, and figure
construction live in the scripts under `R/` and are called from this
document via `source()`.

## Preliminary graphics

```{r}
#| label: fig-logscale
#| fig-cap: |
#|   Daily mean nitrogen dioxide is strongly right-skewed on its original scale
#|   and approximately symmetric after a log transformation.
#| fig-height: 4
analysis_data |>
  plot_log_justification(
    outcome = "no2_mean",
    label = "nitrogen dioxide"
  )
```

```{r}
#| label: fig-trend
#| fig-cap: |
#|   Annual mean concentrations across the selected monitoring locations.
#|   Nitrogen dioxide falls steeply over the study period while ozone drifts
#|   slightly upward, motivating year fixed effects.
#| fig-height: 4.2
analysis_data |>
  summarize_annual_means(outcomes = outcome_columns) |>
  plot_annual_trend()
```

Nitrogen dioxide falls by roughly half across the study period, consistent
with tightening vehicle emissions standards, while ozone is close to flat.
Because these long-run movements are large relative to any weekly effect,
the year term is essential.

# Results

## The weekly cycle

```{r}
#| label: fig-main
#| fig-cap: |
#|   Model-adjusted percentage difference in concentration by day of the week,
#|   relative to Wednesday, with 95% confidence intervals. Estimates control
#|   for month, year, and monitoring location. Nitrogen dioxide falls sharply
#|   across the weekend while ozone rises.
#| fig-height: 5.2
plot_weekly_profile(day_effects)
```

@fig-main is the central result. Nitrogen dioxide is flat from Tuesday
through Friday, dips slightly on Monday, and then falls steeply: about 17%
below Wednesday on Saturday and about 29% below on Sunday. Ozone traces the
mirror image. It is flat midweek and then rises, reaching about 9% above
Wednesday on Saturday and 13% above on Sunday. The confidence intervals are
narrow enough that the weekday and weekend estimates do not overlap for
either pollutant.

## Estimated weekend effect

```{r}
#| label: tbl-effects
#| tbl-cap: |
#|   Estimated weekend effect from the binary specification. Coefficients are
#|   on the log scale; the final columns convert them to percentage
#|   differences.
weekend_effects |>
  format_effect_table() |>
  kable(align = "lrrrrrr")
```

@tbl-effects reports the binary specification. Weekend nitrogen dioxide is
`r round(abs(weekend_effects$pct_change[1]), 1)`% lower than weekday
nitrogen dioxide, with a 95% confidence interval running from
`r round(abs(weekend_effects$pct_high[1]), 1)`% to
`r round(abs(weekend_effects$pct_low[1]), 1)`% lower. Weekend ozone is
`r round(weekend_effects$pct_change[2], 1)`% higher, with an interval
running from `r round(weekend_effects$pct_low[2], 1)`% to
`r round(weekend_effects$pct_high[2], 1)`% higher. Both p-values fall below
any conventional threshold, so we reject both null hypotheses in favour of
the directional alternatives stated in the introduction.

```{r}
#| label: tbl-fit
#| tbl-cap: "Model fit statistics for the binary weekend specification."
weekend_models |>
  summarize_model_fit() |>
  kable(
    col.names = c(
      "Pollutant", "Observations", "Parameters",
      "Adjusted R-squared", "Residual SE"
    ),
    digits = 3,
    format.args = list(big.mark = ",")
  )
```

The models explain a little over half the variation in log NO~2~ and about
42% of the variation in log O~3~. Most of that explanatory power comes from
the location and month terms rather than the weekend indicator, which is
expected: the weekly cycle is a real but modest component of day-to-day
variation, most of which is driven by weather and geography.

## Validity checks {#sec-validity}

```{r}
#| label: fig-resid
#| fig-cap: |
#|   Residuals against fitted values for the nitrogen dioxide model, on a
#|   random sample of 5,000 monitor-days. The spread is broadly even across the
#|   fitted range, supporting the constant-variance assumption.
#| fig-height: 4
#| fig-width: 6.5
weekend_models[["Nitrogen dioxide (NO2)"]] |>
  sample_model_diagnostics() |>
  plot_residual_fitted(label = "nitrogen dioxide")
```

```{r}
#| label: fig-qq
#| fig-cap: |
#|   Normal quantile-quantile plot for the nitrogen dioxide model. The lower
#|   tail departs from the reference line, indicating residuals that are left-
#|   skewed and heavier-tailed than the normal distribution.
#| fig-height: 4
#| fig-width: 6.5
weekend_models[["Nitrogen dioxide (NO2)"]] |>
  sample_model_diagnostics() |>
  plot_residual_qq(label = "nitrogen dioxide")
```

The residual plot in @fig-resid shows no funnel shape and no systematic
curvature, so linearity on the log scale and constant variance are both
reasonable. The quantile-quantile plot in @fig-qq is less flattering: the
residuals are left-skewed and heavy-tailed, so the normality assumption
fails. This matters less than it might appear. With more than 200,000
observations, the sampling distribution of the coefficients is governed by
the central limit theorem rather than by the shape of individual residuals,
so the standard errors and confidence intervals remain approximately valid.
The p-values should nonetheless be read as approximate rather than exact.

A more serious issue is independence, which the design clearly violates:
consecutive days at the same monitor are correlated, because weather
persists. Positive serial correlation of this kind leaves the coefficients
unbiased but makes the reported standard errors too small. The estimated
effects are large enough that this is unlikely to overturn the conclusions,
but the intervals in @tbl-effects should be treated as a lower bound on the
true uncertainty.

# Discussion

## Main findings

Across 24 years and `r n_locations` urban monitoring locations, the weekly
rhythm of human activity is clearly visible in the air. Nitrogen dioxide
falls about 22% on weekends, and the day-by-day profile in @fig-main shows
that the decline tracks the working week almost exactly: flat from Tuesday
to Friday, lower on Saturday, and lowest on Sunday. This is what a
traffic-driven pollutant should look like, and it supports the
traffic-marker hypothesis.

Ozone moves the other way, rising about 10% on weekends. The data therefore
support the titration hypothesis as well. Both null hypotheses are rejected,
and the opposite signs are the substantive finding: cutting emissions does
not lower every pollutant, because the pollutant that harms people at ground
level is partly destroyed by the very emissions that cause the rest of the
problem.

## Why this may or may not be meaningful

The result is meaningful mainly because it is a caution against a natural
intuition. If a city reduced weekday traffic to weekend levels, these
estimates suggest it would enjoy substantially cleaner air in terms of
nitrogen dioxide but might see ozone rise, at least in the urban core where
NO~x~ scavenging dominates. That trade-off is well established in
atmospheric chemistry, and recovering it from a simple regression on public
monitoring data is a reassuring sign that the estimates are picking up real
chemistry rather than noise.

There are reasons for caution. The comparison is observational, not
experimental. "Weekend" is a proxy for reduced commuting traffic, but
weekends differ from weekdays in other ways too, including industrial
activity, heavy goods traffic, and recreational travel, so the coefficient
captures the whole bundle rather than traffic alone. The analysis also says
nothing about *where* the ozone increase matters most; the titration effect
is strongest in dense urban centres and can reverse downwind.

## Critique and possible extensions

Three limitations stand out. First, as noted in @sec-validity, the standard
errors ignore serial correlation within monitoring sites and are therefore
too small; clustering the standard errors by location, or aggregating to
weekly averages, would give more honest intervals. Second, the analysis
controls for season and year with fixed effects but includes no direct
meteorological variables. Temperature, sunlight, and wind speed are the main
drivers of ozone formation, and while there is no reason for weather to
differ systematically between weekdays and weekends, adding weather data
would tighten the model considerably. Third, restricting to the
`r n_locations` best-covered locations buys a balanced panel at the cost of
national representativeness; these are dense urban monitors, exactly the
places where NO~x~ titration is strongest, so the ozone result in particular
should not be generalised to rural areas.

An incidental finding is worth recording. The long-run trend in @fig-trend
shows nitrogen dioxide falling by roughly half between 2000 and 2023 while
ozone stays flat or drifts slightly upward. That is the same trade-off as
the weekly result, playing out across decades rather than days, and it would
make a natural extension: testing whether the weekend ozone effect has grown
or shrunk as background NO~x~ has declined.

# Appendix

```{r}
#| label: tbl-appendix-days
#| tbl-cap: |
#|   Full day-of-week estimates for both pollutants, relative to Wednesday.
day_effects |>
  transmute(
    Pollutant = pollutant,
    Day = term,
    `Change (%)` = round(pct_change, 2),
    `95% CI (%)` = paste0(
      "[", round(pct_low, 2), ", ", round(pct_high, 2), "]"
    ),
    `p value` = format.pval(p_value, digits = 3, eps = 1e-300)
  ) |>
  kable(align = "llrrr")
```

```{r}
#| label: tbl-appendix-locations
#| tbl-cap: "Monitoring locations retained in the analysis sample."
analysis_data |>
  count(location, name = "Monitor-days") |>
  arrange(desc(`Monitor-days`)) |>
  rename(Location = location) |>
  kable(format.args = list(big.mark = ","))
```
